Community Detection Methods for the Generalized Random Dot Product Graph Model

### 2021-05-21 proposal/slides.Rmd
updated slides

2. Popularity Adjusted Block Model
    * Connecting the PABM to the GRDPG
    * Subspace Clustering for Community Detection

@@

$A \sim BernoulliGraph(P)$ iff

* $P \in [0, 1]^{n \times n}$ describes edge probabilities between pairs of 
vertices.
* $A_{ij} \stackrel{ind}{\sim} Bernoulli(P_{ij})$ for each $i < j$.
* $A_{ji} = A_{ij}$ and $A_{ii} = 0$.

@@

**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
$A \sim GRDPG_{3, 1}(X U)$ for $X$ constructed from $\{\lambda^{(kl)}\}_2$ and 
$U \in \mathbb{O}(4)$

Proof:

@@

$$P = X Y^\top$$

@@

* $p = K (K + 1) / 2$
* $q = K (K - 1) / 2$
* $U$ is orthogonal and predetermined for each $K$
* $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$  
$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
to different communities, $x_i^\top x_j = 0$.

**Remark**: Non-uniqueness of the latent configuration  
$A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ 
$\forall Q \in \mathbb{O}(p, q)$

@@

1. Compute the eigenvectors of $A$ that correspond to the $K (K+1)/2$ most 
positive and $K (K-1) / 2$ most negative eigenvalues to construct $V$.
2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
3. Construct graph $G$ using $B$ as its similarity matrix.
4. Partition $G$ into $K$ disconnected subgraphs (e.g., using edge thresholding 
or spectral clustering). Map each partition to the community labels $1, ..., K$.

* $X$ is block diagonal and $U$ is orthogonal $\implies$ each community 
corresponds to a subspace in $\mathbb{R}^d$.
* Subspace property holds even with linear transformation 
$Q \in \mathbb{O}(p, q)$.
* If $P = V \Lambda V^\top$, then $V$ consists of *orthogonal* subspaces.

1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
subject to $x_i = X_{-i} c$ and $c_i^{(i)} = 0$.
2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$
3. Construct affinity matrix $B = |C| + |C^\top|$

* If $X$ obeys the Subspace Detection Property, then $B$ is sparse such that 
$B_{ij} = 0$ if $i$ and $j$ belong to different communities and $\|c_i\| > 0$.
* Step (1) of SSC typically performed via LASSO: 
$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$

@@

* $P_n$ describe the edge probability matrix of the PABM with 
$n$ vertices
* $A_n \sim BernoulliGraph(P_n)$
* $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
$K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues.

@@

@@

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_k(t) \neq f_l(t)$ 
$\forall k, l \in [K]$. 
2. Sample labels
$Z_1, ..., Z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
3. Sample $T_1, ..., T_n \stackrel{iid}{\sim} D$ with support $[0, 1]$. 
4. Set latent positions $X_i = f_{Z_i}(T_i)$ and 
$X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
5. $A \sim BernoulliGraph(X X^\top)$

@@

* Athreya et al. and Rubin-Delanchy et al.: we can approximate properties of 
the latent configurations via ASE.
* General community detection method: Given $A$, $K$, and $d$ (or $p$ and $q$), 
  1. Use ASE to approximate the latent configuration.
  2. Use the appropriate clustering algorithm for the form of the latent
  configuration (manifolds).

@@

n1 <- 2 ** 7
n2 <- 2 ** 7
n <- n1 + n2

a <- .5

@@

@@

```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
par(mar = rep(1, 4))
P <- data.matrix %*% t(data.matrix)
A <- draw.graph(P)
qgraph::qgraph(A, groups = factor(z))
title('RDPG')
```

>>>>

2. Popularity Adjusted Block Model
    * Connecting the PABM to the GRDPG
    * Subspace Clustering for Community Detection
    * Orthogonal Spectral Clustering

@@

$A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.

$A \sim BernoulliGraph(P)$ iff:

1. $P \in [0, 1]^{n \times n}$ describes edge probabilities between pairs of 
vertices.
2. $A_{ij} \stackrel{ind}{\sim} Bernoulli(P_{ij})$ for each $i < j$.

@@

**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
$A \sim GRDPG_{3, 1}(X U)$ for block diagonal $X$ 
constructed from $\{\lambda^{(kl)}\}_2$ and 
predetermined $U \in \mathbb{O}(4)$.

Proof: Decompose $P$ as follows

@@

$$P = X Y^\top = \begin{bmatrix}
\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix}$$

@@

* $p = K (K + 1) / 2$
* $q = K (K - 1) / 2$
* $U$ is orthogonal and predetermined for each $K$
* $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$  
$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
to different communities, then $x_i^\top x_j = 0$.

**Remark** (non-uniqueness of the latent configuration):  
$A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ 
$\forall Q \in \mathbb{O}(p, q)$

**Corollary**: $X$ is block diagonal by community and 
$U$ is orthogonal $\implies$ 
each community corresponds to a subspace in $\mathbb{R}^{K^2}$.

Subspace property holds even with linear transformation 
$Q \in \mathbb{O}(p, q)$.

1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
subject to $x_i = X c$ and $c_i^{(i)} = 0$.
2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
3. Construct affinity matrix $B = |C| + |C^\top|$.

If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
$B_{ij} = 0$ if $i$ and $j$ belong to different communities and $\|c_i\| > 0$.

Step (1) of SSC typically performed via LASSO: 
$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$

Noroozi et al. observed that the rank of $P$ is $K^2$ 
and the columns of $P$ belonging to each community has rank $K$ 
to justify SSC for the PABM.

$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } A_{\cdot, i} = A c_i
\text{ and } c_i^{(i)} = 0$$

They were able to show that this obeys SDP if we replace $A$ with $P$.

GRDPG-based approach: Apply SSC to the ASE of $A$.

$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } 
\hat{x}_i = \hat{X} c_i \text{ and } c_i^{(i)} = 0$$
$$A \approx \hat{X} I_{p, q} \hat{X}^\top$$

Wang and Xu (2016): If data matrix $X$ consists of points lying close to 
low-dimensional subspaces such that:

1. Each point's distance to its subspace is sufficiently small.
2. The cosine of the angles between pairs of subspaces is sufficiently small.
3. The points corresponding to each subspace cover a sufficient amount of 
that subspace.

Then SDP holds with probability 1.

@@

@@

* $P_n$ describe the edge probability matrix of the PABM with $n$ vertices, and 
$A_n \sim BernoulliGraph(P_n)$.
* $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
$K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues.

@@

1. Let $V$ be the eigenvectors of $A$ corresponding to the $K (K+1)/2$ most 
positive and $K (K-1) / 2$ most negative eigenvalues.
2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
3. Construct graph $G$ using $B$ as its similarity matrix.
4. Partition $G$ into $K$ disconnected subgraphs.

@@

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1]^r \mapsto \mathbb{R}^d$ and $f_k(t) \neq f_l(t)$ 
$\forall k, l \in [K]$. 
2. Sample labels
$Z_1, ..., Z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
3. Sample $T_1, ..., T_n \stackrel{iid}{\sim} D$ with support $[0, 1]^r$. 
4. Set latent positions $X_i = f_{Z_i}(T_i)$ and 
$X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
5. $A \sim BernoulliGraph(X X^\top)$

@@

Athreya et al. and Rubin-Delanchy et al.: we can approximate properties of 
the latent configurations via ASE.

General community detection method:  
Given $A$, $K$, and $d$ (or $p$ and $q$),

  1. Use ASE to approximate the latent configuration.
  2. Use the appropriate clustering algorithm for the form of the latent
  configuration (manifolds).

@@

n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2

a <- .1

@@

```{r}
n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
u1 <- runif(n1)
u2 <- runif(n2)
x1 <- cos(pi / 3 * u1)
y1 <- sin(pi / 3 * u1)
x2 <- 1 + cos(pi / 3 * u2 + pi)
y2 <- 1 + sin(pi / 3 * u2 + pi)
data.matrix <- cbind(c(x1, x2), c(y1, y2))
```

@@

```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
par(mar = rep(1, 4))
P <- data.matrix %*% t(data.matrix)
A <- draw.graph(P)
qgraph::qgraph(A, groups = factor(z), layout = 'spring')
title('RDPG')
```
